Generalized Bradley-Terry models

### 2025-03-14 inst/jss_hyper3_real.Rnw
various typos spotted by Gavin

Further, one might follow~\cite{hankin2017_rnw} and consider teams of
players.  In doubles tennis, for example, $\left\lbrace
p_1,p_2\right\rbrace$ might play $\left\lbrace p_3,p_4\right\rbrace$
and win with probability
$\frac{p_1+p_2}{p_1+p_2+p_3+p_4}$~\citep{hankin2010}.  It is natural
to combine this idea with that of a multiplicative factor.  For
example, suppose $p_1$ has a special move against $p_3$, and $p_4$ has
just been given a performance-enhancing drug.  If so, we might have
likelihood function

@@

where $w_{12}$ and $w_{34}$ are the number of wins by each team
respectively.  Then $\lambda$ would represent the effectiveness of
$p_1$'s special move, and $\mu$ the enhancement afforded to $p_4$ by
the drug; values $>1$ would correspond to a positive effect.  In this
paper I generalize the {\tt hyper2} R package~\citep{hankin2017_rnw},
available on CRAN~\citep{rcore2025}, for Bradley-Terry likelihoods to
accommodate this class of multiplicative factors.  The most general
likelihood function would be proportional to

@@

\begin{table}
\centering
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{wins}} &
\multicolumn{4}{c@{}}{\small{Plays away, loses}}\\
\cmidrule(l){2-5}
&   Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers      &   NA &    81  &    69   &      15\\
   Celtic       &   76 &    NA  &    61   &      15\\
   Falkirk      &   17 &    17  &    NA   &       6\\
   Livingston   &    2 &     2  &     3   &      NA\\
\end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{draws}} &
\multicolumn{4}{c@{}}{\small{Plays away, draws}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &       NA  &   44 &     14   &       2\\
   Celtic      &       47  &   NA &     10   &       3\\
   Falkirk     &       12  &   19 &     NA   &       4\\
   Livingston  &       4   &   4  &     4    &      NA\\
   \end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{loses}} &
\multicolumn{4}{c@{}}{\small{Plays away, wins}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &   NA  &  42  &    2  &       0\\
   Celtic      &   46  &  NA  &    8  &       0\\
   Falkirk     &   57  &  43  &   NA  &       8\\
   Livingston  &   12  &  11  &    9  &      NA\\
  \end{tabular}
\caption{The \label{oldfirm} dominance of the Old Firm: 669 matches
  between Rangers, Celtic, Falkirk and Livingston arranged by home
  wins, draws, and home losses.  Data from {\tt soccerworld}}
\end{table}

\begin{equation}\label{firstfewRCFL}
  \left(\frac{\lambda p_R}{\lambda p_R + p_C}\right)^{81}\cdot
  \left(\frac{\lambda p_R}{\lambda p_R + p_F}\right)^{69}\cdots
  \left(\frac{\lambda p_L}{\lambda p_L + p_F}\right)^{81}\cdots
  \left(\frac{p_C}{p_C + \lambda p_R}\right)^{42}\cdots
    \left(\frac{p_F}{p_F + \lambda p_L}\right)^{9}
\end{equation}

@@

\begin{table}
\centering
\begin{tabular}{l|c|c}\\&
{$i$ plays at home} &
{$i$ plays away} \\
  \hline
  $\Prob{(\mbox{$i$ wins})}$                             &
  $\frac{\lambda p_i}{\lambda p_i + D(p_i + p_j) + p_j}$  &
  $\frac{p_i}{p_i + D(p_i + \lambda p_j) + p_j}$          \\
  $\Prob{(\mbox{draw})}$                                 &
  $\frac{D(p_i + p_j)}{\lambda p_i+D(p_i+p_j) + p_j}$     &
  $\frac{D(p_i + p_j)}{p_i + D(p_i+p_j) + \lambda p_j}$   \\
  $\Prob{(\mbox{$i$ loses})}$                            &
  $\frac{p_j}{\lambda p_i + D(p_i+p_j) + p_j}$            &
  $\frac{\lambda p_j}{p_i + D(p_i + p_j) + \lambda p_j}$  \\
\end{tabular}
\caption{Probabilities of a win, draw, loss for the two teams under a
  two-parameter multiplicative Bradley-Terry model\label{windrawloss}}
\end{table}

@@

\begin{figure}[htbp]
  \begin{center}
<<docontourplot, cache=FALSE>>=
contour(lambda, D, matrix(LL, n, n),levels = -2*(1:9), xlab=expression(lambda),ylab=expression(D))
points(maxv[1], maxv[2], pch=16, col='red')
@
\caption{Contours of support\label{contourlambdaD} evaluated over
  a grid.  Maximum likelihood estimate for $(\lambda, D)$ shown as
  a red dot at the evaluate}
  \end{center}
\end{figure}

@@

Further, we may calculate a support function over the
$(\lambda,D)$-plane, shown in figure~\ref{contourlambdaD}.  For
example, we may test the hypothesis that $\lambda=1$ [corresponding to
  the right hand side of figure~\ref{contourlambdaD}] for which the
support is about \Sexpr{round(f(maxv) - jj$objective,2)}, far
exceeding the two units of support per degree of freedom
criteron~\citep{edwards1992}.  Alternatively, we can observe that this
lies in the tail region of its asymptotic null $\chi^2_2$ distribution
with a $p$-value of about \Sexpr{signif(pchisq(f(maxv) -
  jj$objective,lower.tail=FALSE,df=2),digits=2)}.

@@

\begin{equation}
\frac{p_d}{p_a + p_b + p_c + p_d + p_e}\cdot
\frac{p_a}{p_a + p_b + p_c + \mu p_e}\cdot
\frac{\lambda p_c}{\lambda p_b + \lambda p_c + p_e}\cdot
\frac{\lambda p_b}{\lambda p_b + \mu p_e}
\end{equation}

@@

The probability model of Figure~\ref{tikzabcde} is applied, in a
simple one-parameter form, to two datasets: one a set of preferences
of different types of sushi, and one in the field of e-sports.

@@

Here I sugest an alternative in the form of a generalization of
Plackett-Luce likelihood functions for ranking options, using the
multiplicative generalization to Bradley-Terry strengths presented
above.  If we ask our respondent to rank his options, it is highly
probable that he will put RB and BB consecutively (because they are
essentially indistinguishable).  Can we quantify the strength of this
effect?  To do this, we define a bespoke function \code{RB\_BB\_LF()}
which returns a \code{hyper3} log-likelihood function corresponding to
repeated observations of our commuter's reported ranks for the five
options:

@@

Here I analyse a dataset taken from recent e-sports tournament {\em
  ELEAGUE Major: Boston 2018}.  This Valve-sponsored Major featured 24
elite teams competing for a million dollar prize pool; peak concurrent
viewership exceeded $1.3\times 10^6$, the highest in {\em CS:GO}
history at the time.  I study the performance of tournament champions
{\em Cloud9} who, as underdogs, narrowly defeated aggressive European
rivals {\em FaZe Clan} in the final.

>>>>

Further, one might follow~\cite{hankin2017_rnw} and consider teams of
players.  In doubles tennis, for example, $\left\lbrace
p_1,p_2\right\rbrace$ might play $\left\lbrace p_3,p_4\right\rbrace$
and win with probability
$\frac{p_1+p_2}{p_1+p_2+p_3+p_4}$~\citep{hankin2010}.  It is natural
to combine this idea with that of a multiplicative factor.  For
example, suppose $p_1$ has just learned a new special technique, and
$p_4$ has just been given a performance-enhancing drug.  If so, we
might have likelihood function

@@

where $w_{12}$ and $w_{34}$ are the number of wins by each team
respectively.  Then $\lambda$ would represent the effectiveness of
$p_1$'s special technique, and $\mu$ the enhancement afforded to $p_4$
by the drug; values $>1$ would correspond to a positive effect.  In
this paper I generalize the {\tt hyper2} R
package~\citep{hankin2017_rnw}, available on CRAN~\citep{rcore2025},
for Bradley-Terry likelihoods to accommodate this class of
multiplicative factors.  The most general likelihood function would be
proportional to

@@

\begin{table}
\centering
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{wins}} &
\multicolumn{4}{c@{}}{\small{Plays away, loses}}\\
\cmidrule(l){2-5}
&   Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers      &   NA &    81  &    69   &      15\\
   Celtic       &   76 &    NA  &    61   &      15\\
   Falkirk      &   17 &    17  &    NA   &       6\\
   Livingston   &    2 &     2  &     3   &      NA\\
\end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{draws}} &
\multicolumn{4}{c@{}}{\small{Plays away, draws}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &       NA  &   44 &     14   &       2\\
   Celtic      &       47  &   NA &     10   &       3\\
   Falkirk     &       12  &   19 &     NA   &       4\\
   Livingston  &       4   &   4  &     4    &      NA\\
   \end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{loses}} &
\multicolumn{4}{c@{}}{\small{Plays away, wins}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &   NA  &  42  &    2  &       0\\
   Celtic      &   46  &  NA  &    8  &       0\\
   Falkirk     &   57  &  43  &   NA  &       8\\
   Livingston  &   12  &  11  &    9  &      NA\\
  \end{tabular}
\caption{The \label{oldfirm} dominance of the Old Firm: 769 matches
  between Rangers, Celtic, Falkirk and Livingston arranged by home
  wins, draws, and home losses.  Data from {\tt soccerworld}}
\end{table}

\begin{equation}\label{firstfewRCFL}
  \left(\frac{\lambda p_R}{\lambda p_R + p_C}\right)^{81}\cdot
  \left(\frac{\lambda p_R}{\lambda p_R + p_F}\right)^{69}\cdots
  \left(\frac{\lambda p_L}{\lambda p_L + p_F}\right)^{3}\cdots
  \left(\frac{p_C}{p_C + \lambda p_R}\right)^{42}\cdots
    \left(\frac{p_F}{p_F + \lambda p_L}\right)^{9}
\end{equation}

@@

\begin{table}
\centering
\begin{tabular}{l|c|c}\\&
{$i$ plays at home} &
{$i$ plays away} \\
  \hline
  $\Prob{(\mbox{$i$ wins})}$                             &
  $\frac{\lambda p_i}{\lambda p_i + D(p_i + p_j) + p_j}$  &
  $\frac{p_i}{p_i + D(p_i + p_j) + p_j}$          \\
  $\Prob{(\mbox{draw})}$                                 &
  $\frac{D(p_i + p_j)}{\lambda p_i+D(p_i+p_j) + p_j}$     &
  $\frac{D(p_i + p_j)}{p_i + D(p_i+p_j) + \lambda p_j}$   \\
  $\Prob{(\mbox{$i$ loses})}$                            &
  $\frac{p_j}{\lambda p_i + D(p_i+p_j) + p_j}$            &
  $\frac{\lambda p_j}{p_i + D(p_i + p_j) + \lambda p_j}$  \\
\end{tabular}
\caption{Probabilities of a win, draw, loss for the two teams under a
  two-parameter multiplicative Bradley-Terry model\label{windrawloss}}
\end{table}

@@

\begin{figure}[htbp]
  \begin{center}
<<docontourplot, cache=FALSE>>=
contour(lambda, D, matrix(LL, n, n), levels = -2*(1:9),
        xlab = expression(lambda), ylab = expression(D))
points(maxv[1], maxv[2], pch = 16, col = 'red')
abline(v=1, col = 'gray')
@
\caption{Contours of support\label{contourlambdaD} evaluated over
  a grid.  Maximum likelihood estimate for $(\lambda, D)$ shown as
  a red dot at the evaluate.  Vertical gray line shows $H_0\colon\lambda=1$}
  \end{center}
\end{figure}

@@

Further, we may calculate a support function over the
$(\lambda,D)$-plane, shown in figure~\ref{contourlambdaD}.  For
example, we may test the hypothesis that $\lambda=1$ [corresponding to
  the left hand side of figure~\ref{contourlambdaD}] for which the
support is about \Sexpr{round(f(maxv) - jj$objective,2)}, far
exceeding the two units of support per degree of freedom
criteron~\citep{edwards1992}.  Alternatively, we can observe that this
lies in the tail region of its asymptotic null $\chi^2_2$ distribution
with a $p$-value of about \Sexpr{signif(pchisq(f(maxv) -
  jj$objective,lower.tail=FALSE,df=2),digits=2)}.

@@

\begin{equation}
\frac{p_d}{p_a + p_b + p_c + p_d + p_e}\cdot
\frac{p_a}{p_a + p_b + p_c + \mu p_e}\cdot
\frac{\lambda p_c}{\lambda p_b + \lambda p_c + \mu p_e}\cdot
\frac{\lambda p_b}{\lambda p_b + \mu p_e}
\end{equation}

@@

The probability model of Figure~\ref{tikzabcde} is applied, in a
simple one-parameter form, to two datasets: one a set of preferences
of different types of sushi, and one in the field of esports.

@@

Here I suggest an alternative in the form of a generalization of
Plackett-Luce likelihood functions for ranking options, using the
multiplicative generalization to Bradley-Terry strengths presented
above.  If we ask our respondent to rank his options, it is highly
probable that he will put RB and BB consecutively (because they are
essentially indistinguishable).  Can we quantify the strength of this
effect?  To do this, we define a bespoke function \code{RB\_BB\_LF()}
which returns a \code{hyper3} log-likelihood function corresponding to
repeated observations of our commuter's reported ranks for the five
options:

@@

Here I analyse a dataset taken from recent esports tournament {\em
  ELEAGUE Major: Boston 2018}.  This Valve-sponsored Major featured 24
elite teams competing for a million dollar prize pool; peak concurrent
viewership exceeded $1.3\times 10^6$, the highest in {\em CS:GO}
history at the time.  I study the performance of tournament champions
{\em Cloud9} who, as underdogs, narrowly defeated aggressive European
rivals {\em FaZe Clan} in the final.
